Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity

### 2017-03-13 manuscript.Rmd
edit manuscript

---
#author: Fran\c{c}ois Michonneau
title: "Data-driven approaches to identifying knowledge gaps in marine invertebrates biodiversity"
output:
  pdf_document #:
    #pandoc_args: ["-F","pandoc-crossref"]
---

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges, detect changes in the composition of communities, and more generally accelerate the documentation of species diversity. This is especially pressing in marine ecosystems where a significant portion of the biodiversity remains to be discovered and described.

Until recently our knowledge of species and their distributions was difficult to synthesize because it was scattered and difficult to access. However, recent advances in digitization and public databases efforts have made widely available the content of natural history collections, publications about species, their geographical distributions, and their genetic data. Given the urgency with which we need to document biodiversity and the limited means available to do so, we need to use and build on this existing knowledge to identify gaps, and to be more efficient at collecting information and understanding our natural world.

The incorporation of molecular data in the study of biodiversity has highlighted the pervasiveness of cryptic species, and generally the fact that species have smaller ranges than their taxonomic identity implied. Notably, the wide adoption of DNA barcoding for the study of marine diversity has clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult). DNA barcoding has also been seen as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling. By matching DNA barcodes of unidentified species to a public database of accurately identified and vouchered specimens, one can greatly speed up the process of species identification by matching them to already identified species. Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.

Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases. Meta-barcoding enables the parallel sequencing of thousands of DNA barcodes for species that occur in environmental samples. This approach opens new research venues as it becomes possible to estimate the composition of species-rich samples that were virtually inaccessible before (e.g., gut content, marine sediment samples, plankton tows). For instance, by sequencing the plankton of coastal areas, it becomes possible to sample the larval stages of many species that would otherwise be difficult to sample. However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.

With the cost of sequencing going down, and the prospect of having an automated sampling process, these technologies will soon be used for ecosystem and biodiversity monitoring. However, to interpret the results and understand how species distributions shift with global change or other perturbations, we cannot only rely on the identity of the species, but we also need to take advantage of existing knowledge about these species that is coming from natural history collections. In any case, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included public databases.

<!-- here there are 3 things: data availability (having good APIs to access the -->
<!-- data); data comprehensiveness (how many species are captured in the -->
<!-- database); data accuracy (can we trust the data that is found in the -->
<!-- databases) -->

Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US. Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from meta-barcoding studies.

@@

We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species. Most samples were collected on the continental shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).

>>>>

---
#author: Fran\c{c}ois Michonneau
title: "Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity"
output:
  pdf_document #:
    #pandoc_args: ["-F","pandoc-crossref"]
---

@@

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, marine ecosystems are much less understood, and what constitutes the baseline of an healthy and resilient system is not fully characterized. Even basic information such as the number of species that live in the sea is not agreed on, and a significant portion of marine biodiversity remains to be discovered and described. There is therefore a pressing need to improve our knowledge of this environment before species go extinct and its functionning is irreversibly altered. We also have an heterogenous understanding of marine species distributions which limits our understanding of population sizes and connectivity, in turn making the modelization of marine resources and ecosystem resilience less accurate. Until recently, this knowledge was difficult to synthesize because it was scattered and difficult to access. However, recent advances in digitization and public databases efforts have made available the content of natural history collections, information about species occurrences, and genetic data. Given the urgency with which we need to document biodiversity and the limited means available to do so, we argue that we need to build on our existing knowledge to identify gaps, and we need to be more efficient at collecting information to improve our understanding of the natural world.

The incorporation of molecular data into the study of biodiversity has highlighted the pervasiveness of cryptic species, and contributed to the recognition that species have smaller ranges than their taxonomic identity had implied. Notably, despite early criticisms, the wide adoption of DNA barcoding to study marine diversity has invigorated taxonomy. It was also suggested as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling. By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up the process of species identification by matching them to already identified species. Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.

<!-- clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult). DNA barcoding has also been seen as -->

Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases. Metabarcoding has been developed to sequence in parallel the "barcoding gene" across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings. However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.

With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in  reference databases.

<!-- here there are 3 things: data availability (having good APIs to access the data); data comprehensiveness (how many species are captured in the  database); data accuracy (can we trust the data that is found in the databases) -->

Our knowledge about species and their distributions is gathered in regional faunistic lists. They are typically assembled by taxonomic experts. They are comprehensive but usually limited in their geographic scope. Natural history collections have more comprehensive geographic coverage, but the quality of the taxonomic identifications can vary.

Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US. Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from metabarcoding studies.

@@

We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the EEZ waters of the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species. Most samples were collected on the continental shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).
